# Reproducible Research: Peer Assessment 1
Submitted by Tucker Doud

The following report outlines in detail the steps required to process  
and analyze activity data from a personal activity monitoring device.  
The device collects data on _number of steps taken_ at 5 minute intervals  
throughout the day. The data consists of two months of data from an  
anonymous individual collected during the months of October and November  
2012. The original unprocessed file can be found [HERE](https://github.com/jtdoud/RepData_PeerAssessment1/tree/master/Data).

## Loading and preprocessing the data
Load the required library and import the activity data. The date  
column is converted to a date format recognizable by R.

```{r}
library(ggplot2)
setwd("~/My Dropbox/Education/ReproducibleResearch/RepData_PeerAssessment1")
data <- read.csv(file="./Data/activity.csv", stringsAsFactors= F)
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?
Compute the mean steps per day using `tapply` and covert to a data frame.  
Plot the data using the ggplot2 package.

```{r StepsPerDay, message=FALSE}
stepPerDay <- tapply(X=data$steps, INDEX=data$date, FUN=sum)
stepPerDay <- as.data.frame(x=stepPerDay)           

ggplot(data=stepPerDay, aes(x=stepPerDay)) + geom_histogram() + 
        xlab(label= "Steps per Day") + ylab(label= "Count") +
        ggtitle(label= "Histogram\n Steps per Day")
```

Compute the *mean* and *median* total number of steps per day.
```{r}
mean(stepPerDay$stepPerDay, na.rm=T)
median(stepPerDay$stepPerDay, na.rm=T)
```

## What is the average daily activity pattern?
Initialize a new data frame called `meanPerInt` that contains the  
computed _mean per interval_ and plot the result.

```{r AvgDailyActivity}
meanPerInt <- aggregate.data.frame(x= data[ , 1], 
                                   by= list(data$interval), 
                                   FUN= mean, na.rm= T)
names(meanPerInt) <- c("interval", "meanSteps") #rename columns

ggplot(data=meanPerInt, aes(x=interval, y=meanSteps)) + geom_line() +
        xlab(label= "Interval") + ylab(label= "Mean Steps") +
        ggtitle(label= "Mean Steps per Interval")
```

Compute the 5 minute interval that contains the maximum number of steps.
```{r}
max(meanPerInt$meanSteps)
```

## Imputing missing values
The raw data contains missing values. Compute number of missing  
observations
```{r}
length(which(complete.cases(data)== F))

```

The next task is to impute the missing values. Since the mean steps per  
interval is already computed in the section above, this can be used to  
fill in the missing values in the original data.

First, add a rounded mean steps per interval to `meanPerInt` since there  
are not half-steps in reality. Call it `meanStepsR` for _rounded_.
```{r}
meanPerInt$meanStepsR <- round(x=meanPerInt$meanSteps, digits=0)
```

Then, make a data subset of the original NA observations and call it  
`iData` for _imputed_ data. `merge` the data to the rounded mean steps  
from the `meanPerInt` data frame.
```{r}
iData <- merge(x=data[which(is.na(data$steps)), ], 
               y=meanPerInt[, c(1,3)])[, -2]
iData <- iData[, c(3,2,1)] #reorder for rbind
names(iData) <- names(data) #rename for rbind
```

Use `rbind` to link the imputed data subset with original data - complete  
observations only - and call it `nData` for _new data_.
```{r}
nData <- rbind(data[which(!is.na(data$steps)), ], iData)
nData <- nData[order(nData$date, nData$interval), ] #reorder
```

Compute the new _imputed_ steps per day using `nData`. Make a histogram  
with the new data.
```{r ImpStepPerDay, message=FALSE}
nStepPerDay <- tapply(X=nData$steps, INDEX=nData$date, FUN=sum)
nStepPerDay <- as.data.frame(x=nStepPerDay) 

ggplot(data=nStepPerDay, aes(x=nStepPerDay)) + geom_histogram() + 
        xlab(label= "Steps per Day") + ylab(label= "Count") +
        ggtitle(label= "Histogram (Imputed Values)\nSteps per Day")
```

Compute the new mean and median using `nData`.
```{r}
mean(nStepPerDay$nStepPerDay)
median(nStepPerDay$nStepPerDay)
```

Note that the new mean and median values from the imputed data differ  
from the original data set by `r mean(stepPerDay$stepPerDay, na.rm=T)-mean(nStepPerDay$nStepPerDay)` and `r median(stepPerDay$stepPerDay, na.rm=T)-median(nStepPerDay$nStepPerDay)` respectively.

## Are there differences in activity patterns between weekdays and weekends?
The next stage of the analysis examines the difference in mean steps per  
interval between weekends and weekdays. First, convert the dates to  
a weekend/weekday factor using the new data with imputed values.

```{r}
nData$weekDay <- as.factor(weekdays(x=nData$date, abbreviate=T))
nData$weekDayRC <- as.factor(ifelse(test=nData$weekDay=="Sat" | 
                                nData$weekDay== "Sun", yes="Weekend", 
                                no="Weekday"))
```

Then using the new data with the imputed values `nData` compute a new  
data frame with the mean steps per interval and call it `nMeanPerInt`. Do  
a faceted plot to compare weekends to weekdays.

```{r FacetPlot}
nMeanPerInt <- aggregate.data.frame(x= nData[ , 1], 
                                   by= list(nData$interval,
                                            nData$weekDayRC), 
                                   FUN= mean, na.rm= T)

names(nMeanPerInt) <- c("interval", "weekDay", "meanSteps") #rename

ggplot(data=nMeanPerInt, aes(x=interval, y=meanSteps)) + geom_line() +
        facet_grid(weekDay ~ .) +
        xlab(label= "Interval") + ylab(label= "Mean Steps") +
        ggtitle(label= "Mean Steps per Interval (faceted)")
```
