# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r}
library(ggplot2)

setwd("~/My Dropbox/Education/ReproducibleResearch/RepData_PeerAssessment1")
data <- read.csv(file="./Data/activity.csv", stringsAsFactors= F)
data$date <- as.Date(data$date)

```


## What is mean total number of steps taken per day?

```{r}
stepPerDay <- tapply(X=data$steps, INDEX=data$date, FUN=sum)
stepPerDay <- as.data.frame.array(x=stepPerDay)           

ggplot(data=stepPerDay, aes(x=stepPerDay)) + geom_histogram() + 
        xlab(label= "Steps per Day") + ylab(label= "Count") +
        ggtitle(label= "Histogram\n Steps per Day")

mean(stepPerDay$stepPerDay, na.rm=T)
median(stepPerDay$stepPerDay, na.rm=T)

```

## What is the average daily activity pattern?

```{r}
meanPerInt <- aggregate.data.frame(x= data[ , 1], 
                                   by= list(data$interval), 
                                   FUN= mean, na.rm= T)
names(meanPerInt) <- c("interval", "meanSteps")

ggplot(data=meanPerInt, aes(x=interval, y=meanSteps)) + geom_line()

max(meanPerInt$meanSteps)

```


## Imputing missing values

Simply fill the missing values in the original data set with the mean   values from new table created in the preceeding step!

Compute number of missing observations
```{r}
length(which(complete.cases(data)== F))

```

add rounded mean step per interval for imputation

```{r}
meanPerInt$meanStepsR <- round(x=meanPerInt$meanSteps, digits=0)
head(meanPerInt)
```

make data subset of NA observations with rounded average taken from  
meanPerInt frame

```{r}
iData <- merge(x=data[which(is.na(data$steps)), ], 
               y=meanPerInt[, c(1,3)])[, -2]
iData <- iData[, c(3,2,1)] #reorder for rbind
names(iData) <- names(data) #rename for rbind
head(iData)
```

rbind imputed data with original complete data

```{r}
nData <- rbind(data[which(!is.na(data$steps)), ], iData)
nData <- nData[order(nData$date, nData$interval), ] #reorder
```

Make Histogram with new data
```{r}
nStepPerDay <- tapply(X=nData$steps, INDEX=nData$date, FUN=sum)
nStepPerDay <- as.data.frame.array(x=nStepPerDay) 

ggplot(data=nStepPerDay, aes(x=nStepPerDay)) + geom_histogram() + 
        xlab(label= "Steps per Day") + ylab(label= "Count") +
        ggtitle(label= "Histogram (Imputed Values)\nSteps per Day")

mean(nStepPerDay$nStepPerDay)
median(nStepPerDay$nStepPerDay)

```

## Are there differences in activity patterns between weekdays and weekends?
